HO unification from object language to meta language

### 2024-03-22 paper/main.tex
less duplicate code

@@

\begin{elpicode}
type fder fsubst -> fm -> fm -> o.
fder S (fapp A) (fapp B) :- map (fder S) A B.
fder S (flam F) (flam G) :- 
  pi x\ fder S x x => fder S (F x) (G x).
fder _ (fcon C) (fcon C).
fder S (fuva N) R :- set? N S T, fder S T R.
fder S (fuva N) (fuva N) :- unset? N S.

type fderef fsubst -> fm -> fm -> o.                 ~($\customlabel{pred:fder}{\rho s}$)~
fderef S T R :- fder S T T', napp T' R.

type napp fm -> fm -> o.
\end{elpicode}

@@

\begin{elpicode}
type deref subst -> tm -> tm -> o.                   ~($\customlabel{pred:deref}{\sigma t}$)~
deref S (app A) (app B) :- map (deref S) A B.
deref S (lam F) (lam G) :- 
  pi x\ deref S x x => deref S (F x) (G x).
deref _ (con C) (con C).
deref S (uva N A) R :- set? N S F, move F A T, deref S T R.
deref S (uva N A) (uva N B) :- unset? N S, map (deref S) A B.
type move assignment -> list tm -> tm -> o.
move (abs Bo)        [H|L] R :- move (Bo H) L R.
move (val A)         []    A :- !.
move (val (uva N A)) L     (uva N X) :- append A L X.
\end{elpicode}

@@

\begin{elpicode}
  ~\PYG{k+kd}{type} \PYG{n+nf}{(\Eo)} \PYG{k+kt}{ftm -> ftm -> o}~.                           ~($\customlabel{pred:fequal}{\Eo}$)~
  fapp A ~\Eo~fapp B :- map ~(\Eo)~ A B.
  flam F ~\Eo~flam G :- pi x\ x ~\Eo~x => F x ~\Eo~G x.
  fcon C ~\Eo~fcon C.
  fuva N ~\Eo~fuva N.
  flam F ~\Eo~T :-                                       ~($\customlabel{clause:eta1}{\eta_l}$)~
    pi x\ beta T [x] (R x), x ~\Eo~x => F x ~\Eo~R x.
  T ~\Eo~flam F :-                                       ~($\customlabel{clause:eta2}{\eta_r}$)~
    pi x\ beta T [x] (R x), x ~\Eo~x => R x ~\Eo~F x.
  fapp [flam X | L] ~\Eo~T :- beta (flam X) L R, R ~\Eo~T. ~($\customlabel{clause:beta1}{\beta_l}$)~
  T ~\Eo~fapp [flam X | L] :- beta (flam X) L R, T ~\Eo~R. ~($\customlabel{clause:beta2}{\beta_r}$)~
\end{elpicode}

\begin{figure}
  \begin{elpicode}
    ~\PYG{k+kd}{type} \PYG{n+nf}{(\Ee)} \PYG{k+kt}{tm -> tm -> o}~.
    app A ~\Ee~fapp B :- map ~(\Ee)~ A B.
    lam F ~\Ee~flam G :- pi x\ x ~\Ee~x => F x ~\Ee~G x.
    con C ~\Ee~fcon C.
    uva N A ~\Ee~fuva N B :- map ~(\Ee)~ A B.
  \end{elpicode}
  \caption{Equal predicate ML}
  \label{code:eq-pred-meta}
  \Description[code:equal-pred-meta]{code:equal-pred-meta}
\end{figure}

@@

@@

\begin{elpicode}
~\PYG{k+kd}{type} \PYG{n+nf}{(\Ue)} \PYG{k+kt}{tm -> tm -> subst -> subst -> o}~.
\end{elpicode}

@@

This predicate has four
arguments, the two terms $t_1$ and $t_2$ to unify, the old substitution map
$\rho_1$, and the new substitution map $\rho_2$, with the invariant $\rho_1
\subseteq \rho_2$. The first three rules unify terms with same rigid heads, and
call the unification relation on the sub-terms. If $t_1$ (resp. $t_2$) is an
assigned variables, $t_1$ is dereferenced to $t_1'$ (resp. $t_2'$) and the
unification is called between $t_1'$ and $t_2$ (resp. $t_1$ and $t_2'$). If both
terms are unification variables, we test that their arguments are in the pattern
fragment, we allocate a new variable $w$ in $\rho_1$ such that $w$ is the
pruning of the arguments of $t_1$ and $t_2$, we assign both $t_1$ and $t_2$ to
$w$ and return the new mapping $\rho_2$ containing all the new variable
assignment. Finally, if only one of the two terms is an unification variable
$v$, after having verified that $v$ does not occur in the other term $t$, we
bind $v$ to $t$ and return the new substitution mapping.

\old

A key property needed in unification is being able to verify if two terms are
equal wrt a given equational theory. This relation allow to compare terms under
a certain substitution mapping, so that any time a variable $v$ is assigned in a
subterm, a dereferencing of $v$ is performed. After variable dereferencing, the
test for equality is continued on the new-created subterm.

The base equality function over terms can be defined as follows:

The solution we are proposing aim to overcome these unification issues by 1)
compiling the terms $t$ and $u$ of the OL into an internal version $t'$ and $u'$
in the ML; 2) unifying $t'$ and $u'$ at the meta level instantiating meta
variables; 3) decompiling the meta variable into terms of the OL; 4) assigning
the variables of the OL with the decompiled version of their corresponding meta
variables. We claim that $t$ and $u$ unify if and only if $t'$ and $u'$ unify
and that the substitution in the object language is the same as the one returned
by the ML. \todo{same or $\supseteq$ or $\subseteq$}

In the following section we explain how we deal with term (de)compilation and
links between unification variables.

>>>>

\input{code/comp_links}

@@

\input{code/fderef}

@@

\input{code/deref}

@@

@@

\input{code/feq}

\begin{elpicode}
  ~\PYG{k+kd}{type} \PYG{n+nf}{(\Ee)} \PYG{k+kt}{tm -> tm -> o}~.
  app A ~\Ee~fapp B :- map ~(\Ee)~ A B.
  lam F ~\Ee~flam G :- pi x\ x ~\Ee~x => F x ~\Ee~G x.
  con C ~\Ee~fcon C.
  uva N A ~\Ee~fuva N B :- map ~(\Ee)~ A B.
\end{elpicode}

@@

\input{code/ue_type}

@@

This predicate has four
arguments, the two terms $t_1$ and $t_2$ to unify, the old substitution map
$\rho_1$, and the new substitution map $\rho_2$, with the invariant $\rho_1
\subseteq \rho_2$.
% The first three rules unify terms with same rigid heads, and
% call the unification relation on the sub-terms. If $t_1$ (resp. $t_2$) is an
% assigned variables, $t_1$ is dereferenced to $t_1'$ (resp. $t_2'$) and the
% unification is called between $t_1'$ and $t_2$ (resp. $t_1$ and $t_2'$). If both
% terms are unification variables, we test that their arguments are in the pattern
% fragment, we allocate a new variable $w$ in $\rho_1$ such that $w$ is the
% pruning of the arguments of $t_1$ and $t_2$, we assign both $t_1$ and $t_2$ to
% $w$ and return the new mapping $\rho_2$ containing all the new variable
% assignment. Finally, if only one of the two terms is an unification variable
% $v$, after having verified that $v$ does not occur in the other term $t$, we
% bind $v$ to $t$ and return the new substitution mapping.

% \old

% A key property needed in unification is being able to verify if two terms are
% equal wrt a given equational theory. This relation allow to compare terms under
% a certain substitution mapping, so that any time a variable $v$ is assigned in a
% subterm, a dereferencing of $v$ is performed. After variable dereferencing, the
% test for equality is continued on the new-created subterm.

% The base equality function over terms can be defined as follows:

% The solution we are proposing aim to overcome these unification issues by 1)
% compiling the terms $t$ and $u$ of the OL into an internal version $t'$ and $u'$
% in the ML; 2) unifying $t'$ and $u'$ at the meta level instantiating meta
% variables; 3) decompiling the meta variable into terms of the OL; 4) assigning
% the variables of the OL with the decompiled version of their corresponding meta
% variables. We claim that $t$ and $u$ unify if and only if $t'$ and $u'$ unify
% and that the substitution in the object language is the same as the one returned
% by the ML. \todo{same or $\supseteq$ or $\subseteq$}

% In the following section we explain how we deal with term (de)compilation and
% links between unification variables.
